---
title: "GML Poisson"
output: html_document
date: "2025-11-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Load data
```{r}
bees <- read.csv("Eulaema.csv")
head(bees)
```

## Initial plotting
```{r}
hist(bees$Eulaema_nigrita)
plot(bees$altitude, bees$Eulaema_nigrita, pch = 16, cex = 0.6)
plot(bees$MAT, bees$Eulaema_nigrita, pch = 16, cex = 0.6)
plot(bees$effort, bees$Eulaema_nigrita, pch = 16, cex = 0.6)
```

## Fit the poisson model 
The model automatically transforms the data with the natrual logratihm because the choosen family is poisson.
```{r}
m = glm(Eulaema_nigrita~effort+altitude+MAT, family = "poisson", data=bees)
summary(m)
```
We can see here that the residual deviance is 16998 and the degress od freedom is 174. So the deviance is much larger than the degrees of freedom which means that the variance increases unproportional to the mean. Therefore we will have to use a negative binomal error distribution instead. 
\
## Negative binomal

```{r}
library(MASS)
library(MuMIn)
m = glm.nb(Eulaema_nigrita~effort+altitude+MAT, data=bees)
summary(m)
# Calculate the pseudo R square
r.squaredGLMM(m)

# Calculate pseudo r square in a different way
print("How far the model is from a perfect model, where 1 is the optimal number: ")
1-m$deviance / m$null.deviance
```
When looking at the estimates. The intercept is when all 3 predictors are 0. The estimated value for effort is 0.477, the exponential of this is 1.611, meaning that when the effort increases by 1 the bee count increases with 61%. For Altidude this number is (exp(0.00175) = 1.00175) 0.2% and for MAT it is (exp(0.0349) = 1.0355) 3.6%. Standard deviation is around 13% for effort, 25% for altitude and 21% for MAT. 

From this model summary we can see that the residual deviance and the degrees of freedom are much closer and thus this model is a better fit.\
\
Side note : When we make a model with several predictors like this, what the model does is that it makes the regression for each one of the predictors while holding the others constant. SO it is not a combined regression for all 3 predicors, but its more liek three models that holds one predictor constant and then makes three predictions. 
## Plot the results
Effort
MAT
Altitude

Install and load packaged that are needed to make predictions and then plot the results.
```{r}
#install.packages("ggeffects")
library(ggeffects)
library(ggplot2)
```

Make a prediction of y (bee counts) from the predictor "effort". This is done while holding the predictors MAT and altitude constant. This ggpredict function does the same as the base R predict function if you use a datagrid as newdata (which you do when you have multiple predictors)\
When the model is fitted the data is automatically log linked for this model using the natural logratim. When using the ggpredict the data is also automatically transformed back to its original scale. 
```{r}
pred_effort <-  ggpredict(m, terms = "effort")
pred_effort
```
### Effort
Make the plot for predicted bee count plotted against effort.
```{r}
# First plot the predicted counts based on the effort level, this is the natural logarithm regression line
ggplot(pred_effort, aes(x = x, y = predicted)) +
  geom_line(size = 1) + # make the plot a line
  # ribbon is the confidence interval. The data used here is the pred_effort, which has a column conf.low and conf.high. Alpha 0.2 gives the transparacy of the ribbon
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
  # Below points fromt the raw data is added to the plot. 
  geom_point(data = bees, aes(x = effort, y = Eulaema_nigrita),
             inherit.aes = FALSE) +
  labs(x = "Effort",
       y = "Predicted bee count") +
  theme(axis.title =element_text(size=20))

```
### MAT
```{r}
# Make the prediction with MAT as preditor and effort and altitude held constant
pred_MAT <-  ggpredict(m, terms = "MAT")
pred_MAT


# Plot the results
# Plot prediction
ggplot(pred_MAT, aes(x = x, y = predicted)) +
  geom_line(size = 1) + # make the plot a line
  # Add confidence interval
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
  # add raw data points 
  geom_point(data = bees, aes(x = MAT, y = Eulaema_nigrita),
             inherit.aes = FALSE) +
  labs(x = "Mean Average Temperature",
       y = "Predicted bee count") +
  theme(axis.title =element_text(size=20))

```
### Altitude

```{r}
# Make the prediction with altitude as predictor and effort and MAT held constant
pred_altitude <-  ggpredict(m, terms = "altitude")
pred_altitude


# Plot the results
# Plot prediction
ggplot(pred_altitude, aes(x = x, y = predicted)) +
  geom_line(size = 1) + # make the plot a line
  # Add confidence interval
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
  # add raw data points 
  geom_point(data = bees, aes(x = altitude, y = Eulaema_nigrita),
             inherit.aes = FALSE) +
  labs(x = "Altitude",
       y = "Predicted bee count") +
  theme(axis.title =element_text(size=20))

```


```{r}


```

